\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-03-18 Articles/2020-QuantumLogic/Source/quantumlogic.tex
First draft from Andrea

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-1}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{MnSymbol}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\urlstyle{same}
\frenchspacing

\newtheorem{prop}[equation]{Proposition}

@@

\title{Notes on classical/quantum logic }
\author{Gabriele Carcassi, Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

\date{\today}

\begin{abstract}
Notes of classical logic in quantum mechanics.
\end{abstract}

@@

\section{Quantum logic: definition and motivation}

@@

The distributivity in quantum mechanics is ``weird'' because (1) it can be true even if none of the elements of the disjunction is true due to the linearity of the Schroendinger equation and (2) quantum observables do not commute.

@@

In this case $p \wedge \hat{q} \nequiv \bot$ and $p \wedge \hat{r} \nequiv \bot$. Therefore the original claim was $p \wedge (\hat{q} \vee \hat{r})$ if and only if $(p \wedge  q) \vee (p \wedge r)$ which is not the classical logic distributive law. Note that, $\hat{q} \vee \hat{r} \equiv \top$ under condition that ``We measure spin $y$ at time $\hat{t}$''.

\subsubsection{Classical example}
\includegraphics[width=\columnwidth]{Balldrop.png}

@@

\section{Classical and quantum parallel}

Classical state: $\rho(x,p)$

Quantum state: $\psi(x)$

Classical observable $f(x,p)$. Average value $<f> = \int f(x, p) \rho(x, p) dx dp$. Eigenstate $f(x, p) \rho(x, p) = f_0 \rho(x, p)$.

Quantum observable $F$. Average value $<F> = \int \psi^\dagger(x) F \psi(x) dx$. Eigenstate $F \psi(x) = f_0 \psi(x)$.

In both cases, average value equal to $f_0$ means that the contributions of all parts of the ensemble combine to the value $f_0$. Eigenstate $f_0$ means all parts of the ensemble contribute the same value $f_0$.

For example, a classical eigenstate of $H = \frac{p^2}{2m} + \frac{1}{2} k x^2$ with eigenvalue $e_0$ is any function $\rho(x,p)$ that is different from zero only on the ellipses with energy $e_0$. That is, all elements of the distribution have the same energy.

The statement ``the observable $f$ of the system is between $f_0$ and $f_1$'' is therefore unclear. It could be taken to mean ``the average of observable $f$ of the system is between $f_0$ and $f_1$'' or ``the observable $f$ for each part of the ensemble is between $f_0$ and $f_1$''. In the first case, we are identifying a value from the real line, using the standard topology and the standard Borel algebra. In the second case, we are identifying a set of possible functions based on their support.

Now suppose that the $U$ and $V$ are two intervals. We have ``the average $<f>$ is in $U \cup V$'' is  equivalent to ``the average $<f>$ is in $U$'' $\vee$ ``the average $<f>$ is in $V$''. The average value must be in and and only one set. We also have ``the support of $\rho$ is $U \cup V$'' is not equivalent to ``the support of $\rho$ is $U$'' $\vee$ ``the support of $\rho$ is $V$''. The dis tribution could span both sets at the same time. But the lattice generated by such sets is still a $\sigma$-algebra, and therefore follows classical logic.

To check/prove: neither algebras is a sub-algebra of the other; both sub-algebras of the norm-induced one.

\section{Math}

In this section, we want to show how Hilbert spaces already are equipped with a mathematical structure that follows classical logic, namely the Borel algebra. We will see that the algebra typically associated with quantum logic is a subset of this structure. We will see that this classical structure contains other statements that are experimentally significant, like statements about the expectation values. Finally, we will show that the algebra of subspaces of classical distributions is also non-distributive in the same way that quantum logic is.

@@

To show that some of these statements are indeed interesting, let us consider the following result:
\begin{prop}
	Let $H$ be an Hilbert space and $\langle \, , \, \rangle$ be its inner product. Let $X$ be a self-adjoint bounded linear operator. Let $F_X : H \to \mathbb{R}$ the map defined by $F_X(\psi) = \langle \psi , X \psi \rangle$. Let $U \subseteq \mathbb{R}$ be a Borel set. Then $F_X^{-1}(U)$ is a Borel set.
\end{prop}

@@

\section{Old notes}

* We can always use the Borel algebra for our statements. This will include(?) all propositions of ``quantum logic''.

It includes (?) also proposition on expectation values, say of position and momentum. Those will behave exactly like ``point particles''. It also includes (?) propositions on the support.

Show that this is similar to classical distributions.

Claim: in quantum mechanics the ``correct'' logical structure the is algebra of subspaces of the Hilbert space. In this algebra (which is a $\cap$-structure) the meet functions differently than in Boolean logic. Therefore it does not follow classical logic.

Resolution: that algebra does not include many statements that are interesting experimentally. For example, it does not include statements about the expectation values which, in the end, are the result of experiments. (Cross-section?) One

Technical view: the state space of quantum states is a separable Hilbert space; it is a Banach space (complete normed space) and admits a norm-induced topology; it is separable, so it admits a countable basis, so the topology is second countable; therefore the state space has a countably generated Borel-algebra. This, like any set theoretic structure, follows Boolean algebra. Each state is a point and state-statements like ``the system is in state $s_1$'' are singletons in the $\sigma$-algebra. State-statements are all incompatible with each other: either we prepared one state or we prepared another.

If a process with an initial and final state (the internal are disregarded), we have two lattices of propositions. The sample space is the Cartesian product, the $\sigma$-algebra is the product. A measurement process will follow this pattern. We can say ``the system was prepared in state $s$'' and ``the system was measure in state $s$'' and there are two different statements. In the product algebra, we can then mix and match statements with the usual Boolean constructs. Both these are analogous in a classical and quantum case. These will naturally result in different lattices, but they will still be Boolean lattices.

What changes are the probability spaces we can construct. In the quantum case, the probability is not in general a probability measure and has to be consistent with the Born rule (i.e. the inner product).

\bibliography{quantumlogic}

\end{document}

>>>>

\documentclass[11pt, executivepaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{amsfonts}
\usepackage{MnSymbol}
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{geometry}
 \geometry{
 a4paper,
 total={155mm,237mm},
 left=28mm,
 top=32mm,
 }
\usepackage{hyperref}
\hypersetup{colorlinks= true, allcolors=blue}
\setcitestyle{aysep={}}

\newtheorem{prop}{Proposition}

@@

\title{\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}}

\author{Gabriele Carcassi\thanks{Physics Department, University of Michigan, Ann Arbor, MI 48109. E-mail: carcassi@umich.edu} \and Christine A. Aidala\thanks{Physics Department, University of Michigan, Ann Arbor, MI 48109. E-mail: caidala@umich.edu} \and Andrea Oldofredi\thanks{Section de Philosophie, Universit\'e de Lausanne, Switzerland. E-mail: Andrea.Oldofredi@unil.ch}}

@@

\begin{abstract}
\center Notes on classical and quantum logic
\vspace{4mm}

\noindent \emph{Keywords}: Quantum Mechanics; Quantum Logic; Classical Logic; 
\end{abstract}
\vspace{5mm}
\clearpage

\tableofcontents
\vspace{5mm}

\section{Introduction}

As soon as Quantum Mechanics (QM) achieved a definite and coherent mathematical formulation, several physicists and philosophers claimed that its formal structure does not conform to the laws of classical propositional calculus, a belief that is widespread still to this day (cf.\ \cite{Giuntini:2002}, \cite{DallaChiara:2004} and \cite{deRonde:2016}). From the thirties, in fact, various modifications of Classical Logic (CL) were advanced in order to faithfully represent the physical content of quantum theory. For instance, in 1931 the Polish philosopher Zygmunt Zawirski proposed to apply \L ukasiewicz's three-valued logic to QM, starting from considerations about the wave-particle duality and Heisenberg's uncertainty principle. In virtue of the latter, moreover, the logical empiricists Moritz Schlick and Philipp Frank claimed that the conjunction of two or more statements may have no meaning in quantum theory, concluding that the classical conjunction is not always valid in the context of quantum physics, as recalled in \cite{Carnap:1966}. In their view, a typical example of invalid conjunction is one attributing a definite value for both position and momentum to a single quantum particle at a given time $t$. Furthermore, in 1933 Fritz Zwicky suggested that QM rejects the law of excluded-middle.\footnote{For historical details on quantum logic see \cite{Jammer:1974}.} However, it s generally recognized that 
the founding text on Quantum Logic (QL) is Birkhoff and von Neumann's essay ``\emph{The Logic of Quantum Mechanics}'' (\cite{vonNeumann:1936}), where the authors provided the standard account of the propositional calculus obeyed by quantum propositions.\footnote{It is worth noting that in his treatise \emph{Mathematische Grundlagen der Quantenmechanik} published in 1932 von Neumann had already stated that from the algebraic structure of quantum theory it is possible to formulate a new propositional calculus, anticipating seminal ideas that would have been fully expressed of his successive work in collaboration with Garrett Birkhoff.}

A few years after the publication of this paper, the research in quantum logic had a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct formal structures to represent the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now widespread among experts as exemplified e.g.\ by \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it is an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and Classical Mechanics (CM), following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First order and higher order logics will not be discussed here as well as the logic of relativistic quantum theories.}

Recently one of us asked in another essay whether quantum physics necessarily implies a revision of classical logic (cf.\ \cite{Oldofredi:2020}). Alternatively stated, it was asked whether classical logic must be abandoned in the quantum realm or there are quantum theories compatible with a classical propositional calculus. Analyzing Bohmian Mechanics, a well-known interpretation of QM, it has been showed that quantum physics does not necessarily involves a departure from classical logic. More precisely, it has been argued that in the context of BM not only the logical connectives retain their classical meanings---in particular the negation and the disjunction do follow CL---but also the distributivity law holds, contrary to the case of standard QM. This fact is a consequence of the clear ontology of the theory and its its theory of measurement.

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, here we have a three-fold aim: in the first place, we will analyze the main motivations and justifications for the introduction of quantum logic, arguing that they do not entail the failure of CL in the quantum context. In the second place we argue that classical and quantum mechanics share a common algebraic structure which can originate a classical propositional calculus for quantum propositions. A position dissimilar to the received view about QL. Here we will also show that several important statements in quantum mechanics, i.e.\ those concerned with expectation values, are not captured by QL, while they may be taken into account in a classical logical system. Finally, we will argue that it is possible to show that classical propositions may obey a quantum logic. Clearly, we will also spell out the reasons for which CL seems not to work in the context of QM. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: Section \ref{QL} provides a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part), while in Section \ref{Motivations} we review the main motivations that have been given in literature to introduce a quantum logical calculus, explaining also why they do not entail the failure of classical logic in QM. In Section \ref{Math} we provide a mathematical argument aimed at showing not only that QM is compatible with classical logic---sharing a common logical structure with classical mechanics---but also that important sentences of quantum theory are excluded by standard QL, as stated a few lines above.\ In Section \ref{Discussion} we discuss the main philosophical implications of the preceding sections, and Section \ref{conc} concludes the paper.

\section{The Standard Approach to Quantum Logic}
\label{QL}

In order to present the standard approach to QL, let us say a few words about the relation between classical logic and classical mechanics. It is a well known fact that classical propositional logic is equivalent to a Boolean algebra where propositions (atomic and complex) have truth values ``true'' and ``false'' (or 1 and 0), and the main operations among them are conjunction ``$\wedge$'', disjunction ``$\vee$'' and negation ``$\neg$'', the only unary operation. From these logical connectives one can introduce secondary operations, as for instance the material implication ``$\longrightarrow$'', the exclusive or ``$\oplus$'' and logical equivalence ``$\longleftrightarrow$''. Furthermore, it is important for our discussion to underline that the laws of propositional logic include commutativity, associativity, identity and distributivity with respect to $\wedge$ and $\vee$.

Analogously, the algebraic structure underlying classical mechanics is Boolean as well, being commutative, distributive and associative; in this context the logical operations of conjunction, disjunction and negation are replaced respectively by multiplication, addition and complementation among the variables of the theory. This fact has an interesting physical significance, since the variables associated to magnitudes of physical systems can be added and multiplied together---i.e.\ one can sum and multiply measurement results. Both propositional logic and classical mechanics, thus, generate a complemented Boolean lattice; from this fact it follows that the observable algebra of CM ``is isomorphic to a Boolean algebra of propositions with $(\wedge, \vee, \neg)$'' (\cite{David:2015}, p. 79).
To this regard, Beltrametti claims that 
\begin{quote}
Boolean algebras are algebraic models of classical logic (more specifically of classical propositional calculus) with the algebraic operations of meet ($\wedge$) and join ($\vee$) corresponding to the logical connectives ``and'', ``or'', and the unary relation of orthocomplementation corresponding to the logical negation. The rules and tautologies of classical logic have their counterpart in equations and identities of Boolean algebras (\cite{Beltrametti:2004}, p.\ 341). 
\end{quote}

Therefore, it is possible to formally characterize the state of a physical system via logical propositions which can assume the truth values ``true'' or ``false''---depending whether such statements describe true or false state of affairs concerning the system under consideration (cf.\ \cite{Jaeger:2009}, p. 61)---and to perform logical operations among propositions via the connectives $(\wedge, \vee, \neg)$. It is worth noting that in the context of CM it is always determined whether a system instantiates a given property or not; alternatively stated, every logical proposition about physical systems is either true or false, and hence in CM the principle of semantic bivalence holds (for more details cf.\ \cite{Bub:2007}, p.\ 642, \cite{Giuntini:2002}, p.\ 130).

Taking instead into account the mathematical structure and the physical content of standard QM, it is generally claimed that the classical propositional calculus is no longer appropriate to represent the logic obeyed by quantum propositions, since quantum theory does not share the same algebraic structure of classical mechanics. In particular, the algebra of quantum observables is non-commutative and, from a logical perspective, generates a non-distributive orthocomplemented lattice, which is non-Boolean. 
Referring to this, given the axioms of QM Birkhoff and von Neumann showed how the physical content of the theory does not conform to a classical logical structure. The principal aim of their founding essay of quantum logic was to ``discover what logical structures one may hope to find in physical theories which, like quantum mechanics, do not conform to classical logic'' (\cite{vonNeumann:1936}, p.\ 823). The expression quantum logic in this paper must be clarified, since it refers to a quantum propositional calculus in the form of a ``calculus of linear subspaces with respect to \emph{set products}, \emph{linear sums}, and \emph{orthogonal complements}'' which ``resembles the usual calculus of propositions with respect to \emph{and}, \emph{or}, and \emph{not}'' (\emph{ibid}.), where the logical propositions are associated to measurements, tests on quantum systems.\footnote{Cf.\ \cite{Giuntini:2002} and \cite{DallaChiara:2004} for a systematic introduction to various forms of quantum logics, and to \cite{Engesser:2009} for historical and philosophical discussions on the topic.}

Their analysis begins by defining quantum logical propositions as \emph{experimental propositions}, i.e.\ statements affirming that a certain observable $A$ (or a set of observables) measured on a quantum system $s$ has a given value $a_i$ (or a sequence of values). More precisely, the authors stress that both in classical and quantum mechanics observations of physical systems are given by readings of experimental outcomes ($x_1, \dots, x_n$) of \emph{compatible} measurements ($\mu_1, \dots, \mu_n$). The values ($x_1, \dots, x_n$) are elements of what the authors called the ($x_1, \dots, x_n$)-space, i.e.\ the ``observation-space'' of the system in question, whose elements are all the possible combinations of results of the compatible measurements ($\mu_1, \dots, \mu_n$). Hence, the actual values ($x_1, \dots, x_n$) form a subset of such a space. Birkhoff and von Neumann, then, defined the experimental propositions concerning a physical system as the subsets of the observation space associated with it.

Secondly, the authors underlined that in classical and quantum mechanics the states of physical systems are mathematically represented by points in their state spaces---phase space for the classical case, and Hilbert space $\mathcal{H}$ for the quantum case---which provide the maximal information concerning the system. A point in phase space corresponds to the specification of the position and momentum variables of a certain classical system, whereas in QM the points of $\mathcal{H}$ correspond to wave functions. The authors, then, find a connection between subsets of the observation-space of a system and subsets of its Hilbert space, specifying that quantum experimental propositions are mathematically represented by a closed linear subspace of $\mathcal{H}$; this step is crucial in order to obtain the quantum propositional calculus.\ Alternatively, we can say that quantum mechanical operators correspond to propositions with ``yes/no'' (``true/false'') outcome in a logical system\footnote{Cf.\ also \cite{David:2015}, p.\ 78, where we read that ``An orthogonal projector $\textbf{P}$ onto a linear subspace $P\subset\mathcal{H}$ is indeed the operator associated to an observable that can take only the values 1 (and always 1 if the state $\psi\in P$ is in the subspace $P$) or 0 (and always 0 if the state $\psi\in P^{\perp}$ belongs to the orthogonal subspace to $P$).\ Thus we can consider that measuring the observable $\textbf{P}$ is equivalent to perform a test on the system, or to check the validity of a logical proposition $p$ on the system''.} as underlined by Svozi:
\begin{quote}
Any closed linear subspace of---or, equivalently, any projection operator on---a Hilbert space corresponds to an elementary
proposition. The elementary \emph{true--false} proposition can in English be spelled out explicitly as
\begin{quote}
``The physical system has a property corresponding to the associated closed linear subspace'' (\cite{Svozi:1999}, p.\ 1).
\end{quote}
\end{quote}

Thus, since in quantum mechanics physical systems do not have well-defined values for their properties before the measurement of a given operator (contrary to their classical counterparts), or better that quantum observations do not reveal pre-existing values, we have to stress that such quantum propositions refer always to measurements or to preparation procedures. Hence, the sentence ``the particle $s$ has $x$-spin up'' means either that I have measured the observable $S_x$ and I have found the value $+\hbar/2$ associated with the state ``$x$-spin-up'', or equivalently that I prepared $s$ in the $x$-spin-up state. Thus, one cannot simply say, as in classical mechanics, that a certain system has a certain property without having prepared such a system in a given state or having measured it.\footnote{More on this below.}

Now that we have clarified what quantum propositions are, let us introduce Birkhoff and von Neumann's quantum logic. In order to properly define a propositional calculus for QM, one must define the logical operators for conjunction, disjunction and negation and the notion of logical implication. Following \cite{vonNeumann:1936}, the procedure is rather simple: 
\begin{enumerate}
\item The negation of a proposition $p$ is defined by the authors as follows: ``since all operators of quantum mechanics are Hermitian, the mathematical representative of the negative of any experimental proposition is the orthogonal complement of the mathematical representative of the proposition itself'' (\cite{vonNeumann:1936}, pp.\ 826-827). The orthogonal complement $P^{\perp}$ of the subspace $P$ is the set whose elements are all the vectors orthogonal to the elements of $P$. Such an orthogonal complement satisfies the following property: given a subset $P\subset\mathcal{H}$ and a pure state $\psi$, $\psi(P)=1$ iff $\psi(P^{\perp})=0$ and $\psi(P)=0$ iff $\psi(P^{\perp})=1$. As Dalla Chiara and Giuntini underline, ``$\psi$ assigns to an event probability 1 (0, respectively) iff $\psi$ assigns to the orthocomplement of $P$ [notation adapted] probability 0 (1, respectively). As a consequence, one is dealing with an operation that \emph{inverts} the two extreme probability-values, which naturally correspond to the truth-values \emph{truth} and \emph{falsity} (similarly to the classical truth-table of negation)'' (\cite{Giuntini:2002}, p.\ 132).
\item Concerning the conjunction, Birkhoff and von Neumann notice that one can retain the very same set-theoretical interpretation of the classical conjunction in QL, since the intersection of two closed subspaces $P, Q\subset\mathcal{H}$ is still a closed subspace. Thus, one maintains the usual meaning for the conjunction: the pure state $\psi$ verifies $P\cap Q$ iff $\psi$ verifies both $P$ and $Q$. Thus, quantum logic does not introduce a new logical operator for the conjunction.
\item Contrary to the previous case, the logical operator for disjunction cannot be represented by the set-theoretic union, since the set-theoretical union of the two subspaces $P, Q$ will not be in general a subspace, thus, it is not an experimental proposition. Therefore, in QL one introduces the quantum logical disjunction as the the closed span of the subspaces $P, Q$, which is an experimental proposition. Such a statement corresponds to the ``smallest closed subspace containing both $P$ and $Q$'' (\cite{Bacciagaluppi:2009}, p.\ 55). 
\item The logical implication is defined by set theoretical inclusion: given two experimental propositions $p$ and $q$, 
$p$ implies $q$, means that whenever one predicts $p$ with certainty, one can predict also $q$ with certainty, and this is equivalent to state that $p$ is a subset of $q$. This fact is particularly important since the authors showed that ``it is \emph{algebraically} reasonable to try to correlate physical qualities with subsets of phase-space'' and thus, ``physical qualities attributable to any physical system form a partially ordered system'' (\cite{vonNeumann:1936}, p.\ 828, notation adapted).
\end{enumerate}

In this manner we have defined a propositional calculus for quantum experimental propositions, which generates an orthocomplemented lattice. In what follows we analyze the motivations for such a quantum logic, clarifying why it is generally believed that classical propositional calculus cannot be applied in QM.

\section{Quantum Disjunction and Distributivity Law}
\label{Motivations}

The first remarkable difference between QL and CL is the failure of the distributive law. It is a well-known fact that such a law holds in the propositional calculus underlying classical mechanics in virtue of the commutative algebra of classical observables.
Referring to this, \cite{Giuntini:2002} claim that distributivity fails in quantum logic for two main reasons:
\begin{itemize}
\item the non-commutativity of quantum observables,
\item the peculiar behavior of the quantum disjunction---which may be true although neither of the member is true in virtue of the linearity of the Schr\"odinger Equation (SE)---the other major deviation from classical logic. 
\end{itemize}
\noindent Let us then have a closer look to these motivations for the failure of the distributivity law and more generally of CL in the quantum context.

Contrary to the case of classical mechanics, QM relies on a non-commutative algebraic structure\footnote{For details on the mathematical structure of QM and its physical content the reader may refer e.g.\ to \cite{Sakurai1994}, and \cite{Griffiths:2014}. In this essay, it is assumed that the reader has some familiarity with the standard formalism of QM.}: quantum observables generally do not commute, meaning that for any pair of operators $A,B$ we have $[A,B]=AB-BA\neq0$. From a physical perspective this fact entails that if one performs a measurement of $A$ on a quantum system followed by a measurement of $B$, in general one will obtain different results inverting the order of these observations. Moreover, in virtue of the non-commutative structure of quantum theory, to measure an operator $C=A+B$ is not typically equivalent to measure $A$ and $B$ independently and then add the respective results, since $A,B$ will be in general incompatible (cf.\ \cite{David:2015}, p.\ 77). This algebraic fact of QM constitutes the formal basis to prove the Heisenberg uncertainty relation, a theorem of quantum theory which reflects the operational inability to simultaneously measure the values of incompatible operators with arbitrary precision. Such a result, in turn, is usually interpreted ontologically in the sense that quantum system do not instantiate definite properties in non-measurement situations (cf.\ \cite{Sakurai1994}).

In addition, as we will explain in the remainder of this section, the failure of distributivity in quantum logic is due to another particular feature of this new propositional calculus, namely on the behavior of the quantum disjunction, which is due to the presence of quantum superpositions. Indeed, another remarkable difference between QM  and CM is that quantum systems can be in superposition states as a consequence of the linearity of the Schr\"odinger equation---the fundamental dynamical law of QM---which for a single particle reads: 
\begin{align}
\label{SE}
i\hbar\frac{\partial\psi}{\partial t}=\Big(-\frac{\hbar^2}{2m}\nabla_k^2+V\Big)\psi=H\psi,
\end{align}
\noindent where $H$ represents the Hamiltonian operator, defined as the sum of kinetic and potential energy of the system at hand. More specifically, this algebraic property of \eqref{SE} entails that if two wave functions\footnote{In QM the wave function of a system provides the maximal information available about it.} $\psi_1, \psi_2$ are both possible solutions of the same Sch\"odinger equation, then their linear combination (\emph{superposition}) 
\begin{align}
\label{SSE}
\psi_s=\alpha\psi_{1}+\beta\psi_{2}
\end{align} 
is still a solution of the same SE---$|\alpha|^2,|\beta|^2$ (with $\alpha,\beta\in\mathbb{C}$) represent the probabilities to find the system in $\psi_1, \psi_2$ respectively. Notably, the new superposed state $\psi_s$ is also a consistent representation of the system. Referring to this,  \cite{deRonde:2016} explicitly say that in QL ``differently from the case in classical semantics, a quantum disjunction may be true even if neither of it members is true. This reflects, for example, the case in which we are dealing with a state such as that of a spin $1/2$ system which is in a linear combination of states up and down'', as in the example we are going to consider in order to explain the failure of distributivity in the quantum context.

To claim that in quantum logic conjunction and disjunction are not distributive means that given three propositions $p, q$ and $r$ the logical law
\begin{align}
\label{DL}
p\wedge(q\vee r)\longleftrightarrow(p\wedge q)\vee(p\wedge r)
\end{align}

\noindent does not hold.\footnote{In the context of quantum logic, distributivity must be replaced by a weaker law: $(p\wedge q)\vee(p\wedge r)\longrightarrow(p\wedge(q\vee r)).$}
Let us explain this fact with a simple example taken from \cite{Giuntini:2002}. Suppose to consider a $1/2-$spin particle, let's say an electron and assume that this paritlce has spin up along the $x$ axis. From QM we know that there are only two possible spin states in which a particle can be found in each axis after a spin measurement, namely either in the up or in the down state. Furthermore, since spin operators along different axis do not commute, being incompatible observables---$[S_x, S_y]=[S_x, S_z]=[S_y, S_z]\neq0$---they cannot be simultaneously measured. Then, as a consequence of the Heisenberg uncertainty principle, ``both propositions ``$spin_y$ is up'' and ``$spin_y$ is down'' shall be strongly undetermined. However, the disjunction ``either $spin_y$ is up or $spin_y$ is down'' must be true'' (\cite{Giuntini:2002}, pp.\ 133-134), and the same holds for the case of spin in the $z$ direction. Therefore, if $p$ in \eqref{DL} represents the proposition ``the electron has $x$-spin up'', $q$ the proposition ``the electron has $y$-spin up'' and $r$  the proposition `the electron has $y$-spin down'', it is possible to see that the first part of \eqref{DL} express a true statement---i.e.\ $p\wedge(q\vee r)=1$---whereas $(p\wedge q)$ and $(p\wedge r)$ are both false in virtue of the incompatibility between the spin operators among different axes, implying that the second part of the distributive law asserts a false statement, i.e.\ $(p\wedge q)\vee(p\wedge r)=0$. Consequently, this law fails in the context of quantum theory, since we have $p\wedge(q\vee r)=1\longleftrightarrow(p\wedge q)\vee(p\wedge r)=0$. 
\vspace{2mm}

In what follows we will stress that these motivations for QL are not sufficient to claim that CL is rejected by quantum theory.

@@

The distributivity in quantum mechanics is ``weird'' because (1) it can be true even if none of the elements of the disjunction is true due to the linearity of the Schr\"ondinger equation and (2) quantum observables do not commute.

@@

In this case $p \wedge \hat{q} \not\equiv \bot$ and $p \wedge \hat{r} \not\equiv \bot$. Therefore the original claim was $p \wedge (\hat{q} \vee \hat{r})$ if and only if $(p \wedge  q) \vee (p \wedge r)$ which is not the classical logic distributive law. Note that, $\hat{q} \vee \hat{r} \equiv \top$ under condition that ``We measure spin $y$ at time $\hat{t}$''.

\subsubsection{Classical example}
\begin{center}
\includegraphics[scale=.3]{Balldrop.png}
\end{center}

@@

%\section{Classical and quantum parallel}
%
%Classical state: $\rho(x,p)$
%
%Quantum state: $\psi(x)$
%
%Classical observable $f(x,p)$. Average value $<f> = \int f(x, p) \rho(x, p) dx dp$. Eigenstate $f(x, p) \rho(x, p) = f_0 \rho(x, p)$.
%
%Quantum observable $F$. Average value $<F> = \int \psi^\dagger(x) F \psi(x) dx$. Eigenstate $F \psi(x) = f_0 \psi(x)$.
%
%In both cases, average value equal to $f_0$ means that the contributions of all parts of the ensemble combine to the value $f_0$. Eigenstate $f_0$ means all parts of the ensemble contribute the same value $f_0$.
%
%For example, a classical eigenstate of $H = \frac{p^2}{2m} + \frac{1}{2} k x^2$ with eigenvalue $e_0$ is any function $\rho(x,p)$ that is different from zero only on the ellipses with energy $e_0$. That is, all elements of the distribution have the same energy.
%
%The statement ``the observable $f$ of the system is between $f_0$ and $f_1$'' is therefore unclear. It could be taken to mean ``the average of observable $f$ of the system is between $f_0$ and $f_1$'' or ``the observable $f$ for each part of the ensemble is between $f_0$ and $f_1$''. In the first case, we are identifying a value from the real line, using the standard topology and the standard Borel algebra. In the second case, we are identifying a set of possible functions based on their support.
%
%Now suppose that the $U$ and $V$ are two intervals. We have ``the average $<f>$ is in $U \cup V$'' is  equivalent to ``the average $<f>$ is in $U$'' $\vee$ ``the average $<f>$ is in $V$''. The average value must be in and and only one set. We also have ``the support of $\rho$ is $U \cup V$'' is not equivalent to ``the support of $\rho$ is $U$'' $\vee$ ``the support of $\rho$ is $V$''. The dis tribution could span both sets at the same time. But the lattice generated by such sets is still a $\sigma$-algebra, and therefore follows classical logic.
%
%To check/prove: neither algebras is a sub-algebra of the other; both sub-algebras of the norm-induced one.

\section{Math}
\label{Math}
In this section, we want to show how Hilbert spaces already are equipped with a mathematical structure that follows classical logic, namely the Borel algebra. We will see that the algebra typically associated with quantum logic is a subset of this structure. We will see that this classical structure contains other statements that are experimentally significant, like statements about the expectation values. Finally, we will show that the algebra of subspaces of classical distributions is also non-distributive in the same way that quantum logic is.

@@

To show that some of these statements are indeed interesting, let us consider the following result:

\begin{prop}
	Let $H$ be an Hilbert space and $\langle \, , \, \rangle$ be its inner product. Let $X$ be a self-adjoint bounded linear operator. Let $F_X : H \to \mathbb{R}$ the map defined by $F_X(\psi) = \langle \psi , X \psi \rangle$. Let $U \subseteq \mathbb{R}$ be a Borel set. Then $F_X^{-1}(U)$ is a Borel set.
\end{prop}

@@

\section{Discussion}
\label{Discussion}
%\section{Old notes}
%
%* We can always use the Borel algebra for our statements. This will include(?) all propositions of ``quantum logic''.
%
%It includes (?) also proposition on expectation values, say of position and momentum. Those will behave exactly like ``point particles''. It also includes (?) propositions on the support.
%
%Show that this is similar to classical distributions.
%
%
%Claim: in quantum mechanics the ``correct'' logical structure the is algebra of subspaces of the Hilbert space. In this algebra (which is a $\cap$-structure) the meet functions differently than in Boolean logic. Therefore it does not follow classical logic.
%
%Resolution: that algebra does not include many statements that are interesting experimentally. For example, it does not include statements about the expectation values which, in the end, are the result of experiments. (Cross-section?) One 
%
%
%Technical view: the state space of quantum states is a separable Hilbert space; it is a Banach space (complete normed space) and admits a norm-induced topology; it is separable, so it admits a countable basis, so the topology is second countable; therefore the state space has a countably generated Borel-algebra. This, like any set theoretic structure, follows Boolean algebra. Each state is a point and state-statements like ``the system is in state $s_1$'' are singletons in the $\sigma$-algebra. State-statements are all incompatible with each other: either we prepared one state or we prepared another.
%
%If a process with an initial and final state (the internal are disregarded), we have two lattices of propositions. The sample space is the Cartesian product, the $\sigma$-algebra is the product. A measurement process will follow this pattern. We can say ``the system was prepared in state $s$'' and ``the system was measure in state $s$'' and there are two different statements. In the product algebra, we can then mix and match statements with the usual Boolean constructs. Both these are analogous in a classical and quantum case. These will naturally result in different lattices, but they will still be Boolean lattices.
%
%What changes are the probability spaces we can construct. In the quantum case, the probability is not in general a probability measure and has to be consistent with the Born rule (i.e. the inner product).

\section{Conclusion}
\label{conc}
\vspace{5mm}

\textbf{Acknowledgements:} withheld from draft.
\clearpage

\bibliographystyle{apalike}
\bibliography{bibliography}
\end{document}
